Estimating survival in left filtered data

### 2022-01-13 paper.tex
Add github repo link
Prepare to run

The present project replicates the simulation and bias analyses
presented in Chapter 3 of Izano (2017), embeds the problem in the
non-parametric structural causal approach of Pearl (1995), and applies
the WKM and AWKM estimators to the estimation of digestive-system
cancer-free survival under exposure and no censoring rules in the UAW-GM
Cohort. The \texttt{stremr} package was used to estimate the WKM and
AWKM survival curves in both the simulation and the applied analyses
(Sofrygin, van der Laan, and Neugebauer 2021).

@@

The substantive question of interest was the causal effect of
occupational exposure to MWF on cancer incidence risk. Since
occupational MWF exposure occurs only when individuals are at work, we
defined dynamic exposure regimes that depend on employment status. Under
rule \(a_0\), set \(D(t) = 0\), and set \(A(t) = 0\) while \(N(t) = 1\).
Under rule \(a_1\), set \(D(t) = 0\), and set \(A(t) = 1\) while
\(N(t) = 1\). Under both rules, we prevented censoring by death as if it
were intervenable. The causal effect was defined by contrasting the
survival functions \(S_{a_1}(t) = 1 - \E{Y_{a_1}(t)}\) under rule
\(a_1\) to \(S_{a_0}(t) = 1 - \E{Y_{a_0}(t)}\) that under rule \(a_0\).
Note that this causal estimand was defined over \emph{a priori
counterfactuals} not observable in the real world (Frangakis and Rubin
2002). This approach is standard in epidemiologic studies.

@@

\begin{figure}
\caption{Cancer-free survival over time since hire in five simulation scenarios. The true (discrete) survival curve is represented by the solid lines. The average inverse probability weighted Kaplan-Meier (WKM) survival curve is represented by the dashed-line with short dashes. The average Aalen-filtered inverse probability weighted Kaplan-Meier (AWKM) survival curve is represented by the dashed-line with long dashes. Estimated survival curves were averaged over 250 replicates. Salmon color indicates survival and survival estimates under rule $a_0$ when workers are always unexposed. Cyan color indicates those under rule $a_1$ when workers are always exposed while employed.}
\label{fig:survival}
\begin{center}
\includegraphics{/Users/kevinchen/Documents/2021 Fall/256 - Causal Inference/256-final/reports/resources/results.pdf}
\end{center}
\end{figure}

@@

In the simulation study, we showed that under several scenarios
compatible with our hypothetical causal structure, the AWKM survival
estimator had a smaller bias than the WKM estimator. The bias was
smaller when the cumulative incidence of the outcome was low and at
later follow-up time points. Next, we estimated cancer-free survival in
a real-world context. Using data from two plants participating in the
UAW-GM Cohort study, we followed \(26\,182\) individuals starting from
hire to 35 years after hire for incidence of digestive system cancers
(colon, rectal, esophageal, or stomach). This duration of follow-up
spans a sizable proportion of a usual worker's working lifetime. As in
the simulation, the UAW-GM data were longitudinal data with baseline
covariates, time-varying covariates, and a survival outcome.

The exposures of interest were MWF of three types: straight, soluble,
and synthetic (Byers 2006; F. Mirer 2003; F. E. Mirer 2010). Straight
MWFs are hydrocarbon-based fluids that became widely-used by the 1920s.
They continue to occupy a large portion of the MWF market due to their
simple formulation. In straight MWFs, hydrocarbons of different lengths
are mixed together with other additives to attain different properties.
Straight MWFs contain polycyclic aromatic hydrocarbons, long known to be
carcinogenic (IARC 1973). Soluble oils are water-based oil emulsions
first introduced in response to rising oil prices. They now make up the
largest market share of MWFs (Childers 2006). Soluble MWFs are
vulnerable to microbial contamination, so they contain biocides and
chlorinated chemicals. Soluble MWFs have carcinogenic potential from
both their oil components and their additives, but their high lubricity
makes them the most popular fluid type. Synthetic MWFs have the best
toxicological profile, have no oil, and have a higher resistance to
microbial growth. They were introduced into the MWF market in the second
half of the 20th C., but failed to out-perform soluble MWFs in
industrial metalworking applications. Synthetic MWFs contain biocides,
corrosion inhibitors, and chlorinated compounds, some classified as
carcinogenic by the IARC (IARC 1987).

The outcome of interest was digestive system cancer incidence. There is
little past research linking digestive system cancers to MWF exposures,
but there is some evidence suggesting that straight MWFs cause digestive
system cancers (Izano et al. 2019). Cancer incidence was obtained by
linkage to Surveillance, Epidemiology, and End Results (SEER), which
recorded cancer incidence cases starting on January 1, 1973. The cohort
is comprised of individuals hired between 1938 and 1975. Cancer-free
survival to the start of the registry was a left-filtering process
possibly in the presence of the HWSE, as was the case in simulations.
Over the 35 year follow-up period, vital status was obtained through the
Social Security Administration, the National Death Index, as well as
records provided by the UAW. The exposure rules of interest in the
applied analysis were different than those in the simulation study:
\(a_0\) being exposed with 75\% chance in the years exposure was
observed and \(a_1\) being exposed in the years exposure was observed.
Weights were truncated at 1000. Exposure was binary and defined to be
whether there was exposure above the median level of exposure to
straight, soluble, and synthetic MWFs. Counterfactual survival under
rules \(a_0\) and \(a_1\) were estimated using the WKM and AWKM
estimators. Treatment and censoring mechanisms were estimated using
logistic regression conditional on baseline and time-varying
confounders. These logistic regressions were estimated with
stratification by every two years of follow-up. Baseline confounders
included race, sex, plant, and year of hire. Time-varying confounders
included age, cumulative time off, employment status, and exposure to
the metalworking fluids exposure from past years (3 years in the case of
exposure and 6 years in the case of death due to censoring). Summary
statistics for the full study population and those who experienced
digestive system cancer are presented in Table \ref{tab:popchar}.

@@

\begin{figure}[h]
\caption{Cancer-free survival over time since hire under interventions for exposure to straight, soluble, and synthetic metalworking fluids. The estimated inverse probability weighted Kaplan-Meier (WKM) survival curve is represented by the solid. The estimated Aalen-filtered inverse probability weighted Kaplan-Meier (AWKM) survival curve is represented by the dashed-line. Salmon color indicates survival and survival estimates under rule $a_0$; cyan color indicates those under rule $a_1$.}
\label{fig:gm_survival}
\begin{center}
\includegraphics{/Users/kevinchen/Documents/2021 Fall/256 - Causal Inference/256-final/reports/resources/gm_results.pdf}
\end{center}
\end{figure}

\begin{table}[H]
\centering
\caption{Estimated cancer-free survival time over 35-year follow-up.} 
\label{tab:gm_survival}
\begin{tabular}{llrrr}
  \toprule
Rule & Estimator & Straight & Soluble & Synthetic \\ 
  \midrule
$a_0$ & WKM & 29.9606 & 29.9605 & 29.9626 \\ 
  $a_0$ & AWKM & 29.9337 & 29.9398 & 29.9443 \\ 
   \midrule
$a_1$ & WKM & 29.9600 & 29.9584 & 29.9648 \\ 
  $a_1$ & AWKM & 29.9325 & 29.9360 & 29.9479 \\ 
   \bottomrule
\end{tabular}
\end{table}
\begin{table}[H]
\centering
\caption{Estimated difference in average cancer-free survival time over 35-year follow-up comparing rule $a_1$ exposed above the median level of exposure with probability 1\% to $a_0$ never exposed.} 
\label{tab:gm_differences}
\begin{tabular}{lrrr}
  \toprule
Estimator & Straight & Soluble & Synthetic \\ 
  \midrule
WKM & -0.0006 & -0.0021 & 0.0022 \\ 
  AWKM & -0.0012 & -0.0039 & 0.0037 \\ 
   \bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[H]
\centering
\caption{Summary statistics for the cumulative weights used for the WKM estimator at $t = 1$ and $t = 35$ years since hire for each metalworking fluid type and exposure rule.} 
\label{tab:gm_w}
\begin{tabular}{llrrrrrrr}
  \toprule
MWF type & Rule & $t$ & Minimum & Q1 & Median & Mean & Q3 & Maximum \\ 
  \midrule
Straight & $a_0$ &   1 & 1.00 & 1.00 & 1.00 &   1 & 1.00 &   1 \\ 
   & $a_1$ &   1 & 1.00 & 1.00 & 1.00 &   1 & 1.00 &   1 \\ 
  Soluble & $a_0$ &   1 & 1.00 & 1.00 & 1.00 &   1 & 1.00 &   1 \\ 
   & $a_1$ &   1 & 1.00 & 1.00 & 1.00 &   1 & 1.00 &   1 \\ 
  Synthetic & $a_0$ &   1 & 1.00 & 1.00 & 1.00 &   1 & 1.00 &   1 \\ 
   & $a_1$ &   1 & 1.00 & 1.00 & 1.00 &   1 & 1.00 &   1 \\ 
  Straight & $a_0$ &  35 & 0.00 & 2.25 & 4.71 & $9.7 \times 10^{16}$ & 4622.00 & $7.3 \times 10^{20}$ \\ 
   & $a_1$ &  35 & 0.00 & 2.70 & 13.40 & $1.3 \times 10^{18}$ & 22142.45 & $9.8 \times 10^{21}$ \\ 
  Soluble & $a_0$ &  35 & 0.00 & 4.04 & 858.96 & $5.2 \times 10^{15}$ & 74610.19 & $2.3 \times 10^{19}$ \\ 
   & $a_1$ &  35 & 0.00 & 158.56 & 6685.12 & $2.3 \times 10^{17}$ & 1227858.13 & $1.7 \times 10^{21}$ \\ 
  Synthetic & $a_0$ &  35 & 0.00 & 1.02 & 1.25 & $3.7 \times 10^{19}$ & 8.85 & $3.2 \times 10^{23}$ \\ 
   & $a_1$ &  35 & 0.00 & 1.02 & 1.36 & $2.1 \times 10^{21}$ & 22.63 & $1.8 \times 10^{25}$ \\ 
   \bottomrule
\end{tabular}
\end{table}

\begin{figure}
\caption{Median of the cumulative weights for each year since hire with ribbons delimiting the range of values calculated. Salmon color corresponds to the weights used in the estimation of the survival curve under rule $a_0$; cyan corresponds to those for rule $a_1$.}
\label{fig:gm_weights}
\begin{center}
\includegraphics{/Users/kevinchen/Documents/2021 Fall/256 - Causal Inference/256-final/reports/resources/gm_weights.pdf}
\end{center}
\end{figure}

@@

\leavevmode\hypertarget{ref-Childers_2006}{}%
Childers, Jean. 2006. {``The Chemistry of Metalworking Fluids.''} In
\emph{Metalworking Fluids}. CRC Press.

@@

\leavevmode\hypertarget{ref-Frangakis_2002}{}%
Frangakis, Constantine E, and Donald B Rubin. 2002. {``Principal
Stratification in Causal Inference.''} \emph{Biometrics} 58 (1): 21--29.

@@

\leavevmode\hypertarget{ref-IARC_1973}{}%
IARC. 1973. \emph{IARC Monographs on the Evaluation of Carcinogenic Risk
of the Chemical to Man: Certain Polycyclic Aromatic Hydrocarbons and
Heterocyclic Compounds}. Vol. 3. World Health Organization International
Agency for Research on Cancer.

\leavevmode\hypertarget{ref-IARC_1987}{}%
---------. 1987. \emph{IARC Monographs on the Evaluation of the
Carcinogenic Risk of Chemicals to Humans. Overall Evaluations of
Carcinogenicity: An Updating of IARC Monographs, Volumes 1 to 42}.
\emph{World Health Organization}. Vol. 7. World Health Organization
International Agency for Research on Cancer.

@@

\leavevmode\hypertarget{ref-Pearl_1995}{}%
Pearl, Judea. 1995. {``Causal Diagrams for Empirical Research.''}
\emph{Biometrika} 82 (4): 669--88.

>>>>

The present project replicates the simulation and bias analyses
presented in Chapter 3 of Izano (2017), embeds the problem in the
non-parametric structural causal approach of (\textbf{Pearl\_1995?}),
and applies the WKM and AWKM estimators to the estimation of
digestive-system cancer-free survival under exposure and no censoring
rules in the UAW-GM Cohort. The \texttt{stremr} package was used to
estimate the WKM and AWKM survival curves in both the simulation and the
applied analyses (Sofrygin, van der Laan, and Neugebauer 2021). All code
necessary for replicating the analyses presented here may be found on
\href{https://github.com/kvntchn/gm-delayed-entry.git}{GitHub/kvntchn/gm-delayed-entry}.
Note that running the script for the applied analyses will require
additional permissions.

@@

The substantive question of interest was the causal effect of
occupational exposure to MWF on cancer incidence risk. Since
occupational MWF exposure occurs only when individuals are at work, we
defined dynamic exposure regimes that depend on employment status. Under
rule \(a_0\), set \(D(t) = 0\), and set \(A(t) = 0\) while \(N(t) = 1\).
Under rule \(a_1\), set \(D(t) = 0\), and set \(A(t) = 1\) while
\(N(t) = 1\). Under both rules, we prevented censoring by death as if it
were intervenable. The causal effect was defined by contrasting the
survival functions \(S_{a_1}(t) = 1 - \E{Y_{a_1}(t)}\) under rule
\(a_1\) to \(S_{a_0}(t) = 1 - \E{Y_{a_0}(t)}\) that under rule \(a_0\).
Note that this causal estimand was defined over \emph{a priori
counterfactuals} not observable in the real world
(\textbf{Frangakis\_2002?}). This approach is standard in epidemiologic
studies.

@@

\begin{figure}
\caption{Cancer-free survival over time since hire in five simulation scenarios. The true (discrete) survival curve is represented by the solid lines. The average inverse probability weighted Kaplan-Meier (WKM) survival curve is represented by the dashed-line with short dashes. The average Aalen-filtered inverse probability weighted Kaplan-Meier (AWKM) survival curve is represented by the dashed-line with long dashes. Estimated survival curves were averaged over 250 replicates. Salmon color indicates survival and survival estimates under rule $a_0$ when workers are always unexposed. Cyan color indicates those under rule $a_1$ when workers are always exposed while employed.}
\label{fig:survival}
\begin{center}
\includegraphics{/Users/kevinchen/eisen/gm-delayed-entry/reports/resources/results.pdf}
\end{center}
\end{figure}

@@

In the simulation study, we showed that under several scenarios
compatible with our hypothetical causal structure, the AWKM survival
estimator had a smaller bias than the WKM estimator. The bias was
smaller when the cumulative incidence of the outcome was low and at
later follow-up time points. Next, we estimated cancer-free survival in
a real-world context. Using data from two plants participating in the
UAW-GM Cohort study, we followed \(26\,182\) individuals starting from
hire to 65 years after hire for incidence of digestive system cancers
(colon, rectal, esophageal, or stomach). As in the simulation, the
UAW-GM data were longitudinal data with baseline covariates,
time-varying covariates, and a survival outcome.

The exposures of interest were MWF of three types: straight, soluble,
and synthetic (Byers 2006; F. Mirer 2003; F. E. Mirer 2010). Straight
MWFs are hydrocarbon-based fluids that became widely-used by the 1920s.
They continue to occupy a large portion of the MWF market due to their
simple formulation. In straight MWFs, hydrocarbons of different lengths
are mixed together with other additives to attain different properties.
Straight MWFs contain polycyclic aromatic hydrocarbons, long known to be
carcinogenic (\emph{IARC Monographs on the Evaluation of Carcinogenic
Risk of the Chemical to Man: Certain Polycyclic Aromatic Hydrocarbons
and Heterocyclic Compounds} 1973). Soluble oils are water-based oil
emulsions first introduced in response to rising oil prices. They now
make up the largest market share of MWFs (\textbf{Childers\_2006?}).
Soluble MWFs are vulnerable to microbial contamination, so they contain
biocides and chlorinated chemicals. Soluble MWFs have carcinogenic
potential from both their oil components and their additives, but their
high lubricity makes them the most popular fluid type. Synthetic MWFs
have the best toxicological profile, have no oil, and have a higher
resistance to microbial growth. They were introduced into the MWF market
in the second half of the 20th C., but failed to out-perform soluble
MWFs in industrial metalworking applications. Synthetic MWFs contain
biocides, corrosion inhibitors, and chlorinated compounds, some
classified as carcinogenic by the IARC (\textbf{IARC\_1987?}).

The outcome of interest was digestive system cancer incidence. There is
little past research linking digestive system cancers to MWF exposures,
but there is some evidence suggesting that straight MWFs cause digestive
system cancers (Izano et al. 2019). Cancer incidence was obtained by
linkage to Surveillance, Epidemiology, and End Results (SEER), which
recorded cancer incidence cases starting on January 1, 1973. The cohort
is comprised of individuals hired between 1938 and 1975. Cancer-free
survival to the start of the registry was a left-filtering process
possibly in the presence of the HWSE, as was the case in simulations.
Over the 65 year follow-up period, vital status was obtained through the
Social Security Administration, the National Death Index, as well as
records provided by the UAW. The exposure rules of interest in the
applied analysis were different than those in the simulation study:
\(a_0\) having cumulative exposure of more than 0.05
mg/m\textsuperscript{3} years and \(a_1\) having no cumulative exposure.
Weights were truncated at 1000. Exposure was binary and defined to be
whether there was exposure above the median level of exposure to
straight, soluble, and synthetic MWFs. Counterfactual survival under
rules \(a_0\) and \(a_1\) were estimated using the WKM and AWKM
estimators. Treatment and censoring mechanisms were estimated using
logistic regression conditional on baseline and time-varying
confounders. These logistic regressions were estimated with
stratification by every two years of follow-up. Baseline confounders
included race, sex, plant, and year of hire. Time-varying confounders
included age, cumulative time off, employment status, and exposure to
the metalworking fluids exposure from past years (3 years in the case of
exposure and 6 years in the case of death due to censoring). Summary
statistics for the full study population and those who experienced
digestive system cancer are presented in Table \ref{tab:popchar}.

@@

\begin{figure}[h]
\caption{Cancer-free survival over time since hire under interventions for exposure to straight, soluble, and synthetic metalworking fluids. The estimated inverse probability weighted Kaplan-Meier (WKM) survival curve is represented by the solid. The estimated Aalen-filtered inverse probability weighted Kaplan-Meier (AWKM) survival curve is represented by the dashed-line. Salmon color indicates survival and survival estimates under rule $a_0$; cyan color indicates those under rule $a_1$.}
\label{fig:gm_survival}
\begin{center}
\includegraphics{/Users/kevinchen/eisen/gm-delayed-entry/reports/resources/gm_results.pdf}
\end{center}
\end{figure}

\begin{table}[H]
\centering
\caption{Estimated cancer-free survival time over 65-year follow-up.} 
\label{tab:gm_survival}
\begin{tabular}{llrrrr}
  \toprule
Rule & Estimator & Straight & Soluble & Synthetic & Any \\ 
  \midrule
$a_0$ & WKM & 55.9808 & 56.0093 & 56.2051 & 55.8072 \\ 
  $a_0$ & AWKM & 55.8864 & 55.8952 & 56.1300 & 55.6712 \\ 
   \midrule
$a_1$ & WKM & 56.2258 & 56.2760 & 56.1323 &  \\ 
  $a_1$ & AWKM & 56.0953 & 56.1824 & 56.0066 &  \\ 
   \bottomrule
\end{tabular}
\end{table}
\begin{table}[H]
\centering
\caption{Estimated difference in average cancer-free survival time over 65-year follow-up comparing rule $a_1$ exposed above the median level of exposure with probability 1\% to $a_0$ never exposed.} 
\label{tab:gm_differences}
\begin{tabular}{lrrrr}
  \toprule
Estimator & Straight & Soluble & Synthetic & Any \\ 
  \midrule
WKM & 0.2449 & 0.2667 & -0.0727 &  \\ 
  AWKM & 0.2089 & 0.2872 & -0.1233 &  \\ 
   \bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[H]
\centering
\caption{Summary statistics for the cumulative weights used for the WKM estimator at $t = 1$ and $t = 65$ years since hire for each metalworking fluid type and exposure rule.} 
\label{tab:gm_w}
\begin{tabular}{llrrrrrrr}
  \toprule
MWF type & Rule & $t$ & Minimum & Q1 & Median & Mean & Q3 & Maximum \\ 
  \midrule
Straight & $a_0$ &   1 &   1 &   1 &   1 &   1 &   1 &   1 \\ 
   & $a_1$ &   1 &   1 &   1 &   1 &   1 &   1 &   1 \\ 
  Soluble & $a_0$ &   1 &   1 &   1 &   1 &   1 &   1 &   1 \\ 
   & $a_1$ &   1 &   1 &   1 &   1 &   1 &   1 &   1 \\ 
  Synthetic & $a_0$ &   1 &   1 &   1 &   1 &   1 &   1 &   1 \\ 
   & $a_1$ &   1 &   1 &   1 &   1 &   1 &   1 &   1 \\ 
  Any & $a_0$ &   1 &   1 &   1 &   1 &   1 &   1 &   1 \\ 
   & $a_1$ &   1 &   1 &   1 &   1 &   1 &   1 &   1 \\ 
  Straight & $a_0$ &  57 &   0 & 4.6 & 5.6 &  93 & $2 \times 10^{2}$ & $3.9 \times 10^{2}$ \\ 
   & $a_1$ &  57 &   0 &   0 &   0 & 2.4 & 5.2 & 5.8 \\ 
  Soluble & $a_0$ &  57 &   0 & 3.9 & $1.5 \times 10^{2}$ & $1.7 \times 10^{2}$ & $3.1 \times 10^{2}$ & $3.8 \times 10^{2}$ \\ 
   & $a_1$ &  57 &   0 &   0 &   0 &  41 &   0 & $3.7 \times 10^{2}$ \\ 
  Synthetic & $a_0$ &  57 &   0 & 2.6 & 3.2 &  97 & 3.4 & $8.5 \times 10^{2}$ \\ 
   & $a_1$ &  57 &   0 & 2.5 & 2.9 & 2.4 & 3.3 & 3.6 \\ 
  Any & $a_0$ &  57 &   0 & 3.6 &  15 & $1.4 \times 10^{2}$ & $3 \times 10^{2}$ & $4.1 \times 10^{2}$ \\ 
   & $a_1$ &  57 &   0 &   0 &   0 &   0 &   0 &   0 \\ 
   \bottomrule
\end{tabular}
\end{table}

\begin{figure}
\caption{Median of the cumulative weights for each year since hire with ribbons delimiting the range of values calculated. Salmon color corresponds to the weights used in the estimation of the survival curve under rule $a_0$; cyan corresponds to those for rule $a_1$.}
\label{fig:gm_weights}
\begin{center}
\includegraphics{/Users/kevinchen/eisen/gm-delayed-entry/reports/resources/gm_weights.pdf}
\end{center}
\end{figure}

@@

@@

@@

\leavevmode\hypertarget{ref-IARC_1973}{}%
\emph{IARC Monographs on the Evaluation of Carcinogenic Risk of the
Chemical to Man: Certain Polycyclic Aromatic Hydrocarbons and
Heterocyclic Compounds}. 1973. Vol. 3. World Health Organization
International Agency for Research on Cancer.

@@

### 2022-01-13 reports/04-gm.Rmd
Prepare to run

In the simulation study, we showed that under several scenarios compatible with our hypothetical causal structure, the AWKM survival estimator had a smaller bias than the WKM estimator. The bias was smaller when the cumulative incidence of the outcome was low and at later follow-up time points. Next, we estimated cancer-free survival in a real-world context. Using data from two plants participating in the UAW-GM Cohort study, we followed $26\,182$ individuals starting from hire to 35 years after hire for incidence of digestive system cancers (colon, rectal, esophageal, or stomach). This duration of follow-up spans a sizable proportion of a usual worker's working lifetime. As in the simulation, the UAW-GM data were longitudinal data with baseline covariates, time-varying covariates, and a survival outcome.

@@

The outcome of interest was digestive system cancer incidence. There is little past research linking digestive system cancers to MWF exposures, but there is some evidence suggesting that straight MWFs cause digestive system cancers [@Izano_2019].
<!-- We obtained cancer incidence data for those at Plant 3 via linkage to the Michigan Cancer Registry, which recorded cancer incidence cases starting on January 1, 1985. -->
Cancer incidence was obtained by linkage to Surveillance, Epidemiology, and End Results (SEER), which recorded cancer incidence cases starting on January 1, 1973.
The cohort is comprised of individuals hired between 1938 and 1975. Cancer-free survival to the start of the registry was a left-filtering process possibly in the presence of the HWSE, as was the case in simulations. Over the 35 year follow-up period, vital status was obtained through the Social Security Administration, the National Death Index, as well as records provided by the UAW. The exposure rules of interest in the applied analysis were different than those in the simulation study: $a_0$ being exposed with 75\% chance in the years exposure was observed and $a_1$ being exposed in the years exposure was observed. Weights were truncated at 1000. Exposure was binary and defined to be whether there was exposure above the median level of exposure to straight, soluble, and synthetic MWFs. Counterfactual survival under rules $a_0$ and $a_1$ were estimated using the WKM and AWKM estimators. Treatment and censoring mechanisms were estimated using logistic regression conditional on baseline and time-varying confounders. These logistic regressions were estimated with stratification by every two years of follow-up. Baseline confounders included race, sex, plant, and year of hire. Time-varying confounders included age, cumulative time off, employment status, and exposure to the metalworking fluids exposure from past years (3 years in the case of exposure and 6 years in the case of death due to censoring). Summary statistics for the full study population and those who experienced digestive system cancer are presented in Table \ref{tab:popchar}.

@@

gm_survival.tab <- dcast(
	melt(gm.surv[t <= 30],
			 measure.vars = c("WKM", "AWKM"),
			 variable.name = "Estimator",
			 value.name = "Survival")[,.(
		Survival = sum(Survival)
	), .(MWF = mwf, Rule = rule, Estimator)],
	... ~ MWF,
	value.var = "Survival"
)

@@

gm_differences.tab <- gm_survival.tab[,.(
		Straight  = Straight[2] - Straight[1],
		Soluble   = Soluble[2] - Soluble[1],
		Synthetic = Synthetic[2] - Synthetic[1]
	), .(Estimator)]

@@

```{r}
print(xtable(gm_survival.tab,
			 caption = "Estimated cancer-free survival time over 35-year follow-up.",
			 label = "tab:gm_survival", digits = 4),
			hline.after = c(-1,0,2, nrow(gm_survival.tab)),
			floating = T)

print(xtable(gm_differences.tab,
			 caption = "Estimated difference in average cancer-free survival time over 35-year follow-up comparing rule $a_1$ exposed above the median level of exposure with probability 1\\% to $a_0$ never exposed.",
			 label = "tab:gm_differences", digits = 4),
			floating = T)

@@

```{r}
gm_w.small_tab <- rbindlist(list(
	gm_w.tab[t == 1,],
	gm_w.tab[t == 35,]))

@@

gm_w.small_tab$mwf <- rep(as.vector(sapply(
	c("Straight", "Soluble", "Synthetic"), function(x) {
	c(x, "")})), 2)

@@

print(xtable(
	gm_w.small_tab,
	caption = "Summary statistics for the cumulative weights used for the WKM estimator at $t = 1$ and $t = 35$ years since hire for each metalworking fluid type and exposure rule.",
	label = "tab:gm_w",
	display = c("s", "s", "s", "d", "f", "f", "f", "g", "f", "g")),
	math.style.exponents = TRUE,
	floating = T)

>>>>

In the simulation study, we showed that under several scenarios compatible with our hypothetical causal structure, the AWKM survival estimator had a smaller bias than the WKM estimator. The bias was smaller when the cumulative incidence of the outcome was low and at later follow-up time points. Next, we estimated cancer-free survival in a real-world context. Using data from two plants participating in the UAW-GM Cohort study, we followed $26\,182$ individuals starting from hire to 65 years after hire for incidence of digestive system cancers (colon, rectal, esophageal, or stomach). As in the simulation, the UAW-GM data were longitudinal data with baseline covariates, time-varying covariates, and a survival outcome.

@@

The outcome of interest was digestive system cancer incidence. There is little past research linking digestive system cancers to MWF exposures, but there is some evidence suggesting that straight MWFs cause digestive system cancers [@Izano_2019].
<!-- We obtained cancer incidence data for those at Plant 3 via linkage to the Michigan Cancer Registry, which recorded cancer incidence cases starting on January 1, 1985. -->
Cancer incidence was obtained by linkage to Surveillance, Epidemiology, and End Results (SEER), which recorded cancer incidence cases starting on January 1, 1973.
The cohort is comprised of individuals hired between 1938 and 1975. Cancer-free survival to the start of the registry was a left-filtering process possibly in the presence of the HWSE, as was the case in simulations. Over the 65 year follow-up period, vital status was obtained through the Social Security Administration, the National Death Index, as well as records provided by the UAW. The exposure rules of interest in the applied analysis were different than those in the simulation study: $a_0$ having cumulative exposure of more than 0.05 mg/m^3^ years and $a_1$ having no cumulative exposure. Weights were truncated at 1000. Exposure was binary and defined to be whether there was exposure above the median level of exposure to straight, soluble, and synthetic MWFs. Counterfactual survival under rules $a_0$ and $a_1$ were estimated using the WKM and AWKM estimators. Treatment and censoring mechanisms were estimated using logistic regression conditional on baseline and time-varying confounders. These logistic regressions were estimated with stratification by every two years of follow-up. Baseline confounders included race, sex, plant, and year of hire. Time-varying confounders included age, cumulative time off, employment status, and exposure to the metalworking fluids exposure from past years (3 years in the case of exposure and 6 years in the case of death due to censoring). Summary statistics for the full study population and those who experienced digestive system cancer are presented in Table \ref{tab:popchar}.

@@

gm_survival.tab <- dcast(
	melt(gm.surv[t <= 65],
			 measure.vars = c("WKM", "AWKM"),
			 variable.name = "Estimator",
			 value.name = "Survival")[,.(
		Survival = sum(Survival, na.rm = T)
	), .(MWF = mwf, Rule = rule, Estimator)],
	... ~ MWF,
	value.var = "Survival"
)

@@

gm_differences.tab <- gm_survival.tab[,.(
		Straight  = Straight[2] - Straight[1],
		Soluble   = Soluble[2] - Soluble[1],
		Synthetic = Synthetic[2] - Synthetic[1],
		Any = Any[2] - Any[1]
	), .(Estimator)]

gm_ci.tab <- dcast(
	melt(gm.surv[t == 55],
			 measure.vars = c("WKM", "AWKM"),
			 variable.name = "Estimator",
			 value.name = "Survival")[,.(
		"Cumulative incidence" = 1 - Survival
	), .(MWF = mwf, Rule = rule, Estimator)],
	... ~ MWF,
	value.var = "Cumulative incidence"
)

gm_cir.tab <- gm_ci.tab[,.(
		Straight  = Straight[2]/Straight[1],
		Soluble   = Soluble[2]/Soluble[1],
		Synthetic = Synthetic[2]/Synthetic[1],
		Any = Any[2]/Any[1]
	), .(Estimator)]

pander(gm_cir.tab)
```

@@

```{r}
print(xtable(gm_survival.tab,
			 caption = "Estimated cancer-free survival time over 65-year follow-up.",
			 label = "tab:gm_survival", digits = 4),
			hline.after = c(-1,0,2, nrow(gm_survival.tab)),
			floating = T)

print(xtable(gm_differences.tab,
			 caption = "Estimated difference in average cancer-free survival time over 65-year follow-up comparing rule $a_1$ exposed above the median level of exposure with probability 1\\% to $a_0$ never exposed.",
			 label = "tab:gm_differences", digits = 4),
			floating = T)

@@

```{r}
gm_w.small_tab <- rbindlist(list(
	gm_w.tab[t == 1,],
	gm_w.tab[t == 57,]))

@@

gm_w.small_tab$mwf <- rep(as.vector(sapply(
	c("Straight", "Soluble", "Synthetic", "Any"), function(x) {
	c(x, "")})), 2)

@@

print(xtable(
	gm_w.small_tab,
	caption = "Summary statistics for the cumulative weights used for the WKM estimator at $t = 1$ and $t = 65$ years since hire for each metalworking fluid type and exposure rule.",
	label = "tab:gm_w",
	display = c("s", "s", "s", "d", "g", "g", "g", "g", "g", "g")),
	math.style.exponents = TRUE,
	floating = T)
